Constructive forcing

### 2023-09-14 slides-herrsching2023.tex
Instructive(?) examples for KJ

\title{New modal operators for constructive mathematics}

\author{Ingo Blechschmidt}
\date{May 2nd, 2023}

@@

  \emph{Proof.} (special case) Write~$M = \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
  have~$u, v$ with
  \[
    u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
    \quad\text{and}\quad
    v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
  \]
  Hence
  $
    1 = (vy) (ux) = (uy) (vx) = 0
  $. \qed
\end{frame}}

@@

  We then define~``\hil{$P \mid \sigma$}'' (``$P$ bars~$\sigma$'') inductively by the clauses:
  \begin{enumerate}
    \item If~$P\sigma$, then~$P \mid \sigma$.
    \item If~$P \mid \tau$ for all~$\tau \in R$, where~$R$ is some covering
    of~$\sigma$, then~$P \mid \sigma$.
  \end{enumerate}
  So~$P \mid \sigma$ expresses in a \hil{direct inductive fashion}:
  \[ \text{``No matter
  how~$\sigma$ evolves to a better approximation~$\tau$, eventually~$P\tau$
  will hold.''} \]
  \pause

  We use quantifier-like notation: ``$\nabla(\tau \preccurlyeq \sigma)\_
  P\tau$'' means ``$P \mid \sigma$''.
\end{frame}
% BOARD:
% - examples for P | σ
% - abuse of notation

@@

\newcommand{\defeqvi}{\quad iff\quad}
\begin{frame}{The $\nabla$-translation}
  \small
  For bounded first-order formulas over the (large) first-order signature which has
  \begin{enumerate}
    \scriptsize
    \item one sort~$\underline{X}$ for each set~$X$ in the base universe,
    \\[-1.2em]
    \item one~$n$-ary function symbol~$\underline{f} : \underline{X_1} \times
    \cdots \times \underline{X_n} \to \underline{Y}$ for each map~$f : X_1 \times
    \cdots \times X_n \to Y$,
    \\[-1.2em]
    \item one~$n$-ary relation symbol~$\underline{R} \hookrightarrow
    \underline{X_1} \times \cdots \times \underline{X_n}$ for each relation~$R
    \subseteq X_1 \times \cdots \times X_n$, and
    \\[-1.2em]
    \item an additional unary relation symbol~$G \hookrightarrow \underline{L}$
    (for the \emph{generic filter} of~$L$),
  \end{enumerate}
  we recursively define:
  \scriptsize
  \begin{tabbing}
    \quad \= $\sigma \forces \forces \forall(x\?\underline{X})\_ \varphi$ \=
    \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\
    \forall(x_0 \in X)\_ \tau \forces \varphi[\underline{x_0}/x]$.\qquad\quad \=
    $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
    \= $\sigma \forces \underline{R}(\underline{s_1},\ldots,\underline{s_n})$ \= \defeqvi $s = t$. \= \kill

@@

    \> $\sigma \forces \forall(x\?\underline{X})\_ \varphi$
    \> \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\ \forall(x_0 \in X)\_ \tau \forces
    \varphi[\underline{x_0}/x]$.
    \> $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
    \> \defeqvi $\hil{$\nabla\sigma\_$}\ \exists(x_0 \in X)\_ \sigma \forces \varphi[\underline{x_0}/x]$.
  \end{tabbing}
  \small
  Finally, we say that~$\varphi$ ``holds in~$V^\nabla$'' iff for all~$\sigma
  \in L$, $\sigma \forces \varphi$.
\end{frame}

@@

>>>>

\title{Constructive forcing}

\author{Ingo Blechschmidt}
\date{September 20th to September 16th, 2023}

@@

  \emph{Proof.} (special case) Write~$M = \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right)$. By surjectivity,
  have~$u, v \in A$ with
  \[
    u \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}1\\0\end{smallmatrix}\right)
    \quad\text{and}\quad
    v \left(\begin{smallmatrix}x\\y\end{smallmatrix}\right) = \left(\begin{smallmatrix}0\\1\end{smallmatrix}\right).
  \]
  Hence
  $
    1 = (vy) (ux) = (uy) (vx) = 0
  $. \qed
\end{frame}}

@@

  We then define~``\hil{$P \mid \sigma$}'' (``$P$ bars~$\sigma$'') inductively by the following clauses:
  \begin{enumerate}
    \item If~$P\sigma$, then~$P \mid \sigma$.
    \item If~$P \mid \tau$ for all~$\tau \in R$, where~$R$ is some covering
    of~$\sigma$, then~$P \mid \sigma$.
  \end{enumerate}
  So~$P \mid \sigma$ expresses in a \hil{direct inductive fashion}:
  \[ \text{``No matter
  how~$\sigma$ evolves to a better approximation~$\tau$, eventually~$P\tau$
  will hold.''} \]

  \pause
  We use quantifier-like notation: ``$\nabla(\tau \preccurlyeq \sigma)\_
  P\tau$'' means ``$P \mid \sigma$''.
\end{frame}
% BOARD:
% - examples for P | σ
% - abuse of notation

@@

\newcommand{\defeqvi}{\quad iff\quad}
\begin{frame}{The $\nabla$-translation}
  \small
  \only<1>{For bounded first-order formulas over the (large) first-order signature which has
  \begin{enumerate}
    \scriptsize
    \item one sort~$\underline{X}$ for each set~$X$ in the base universe,
    \\[-1.2em]
    \item one~$n$-ary function symbol~$\underline{f} : \underline{X_1} \times
    \cdots \times \underline{X_n} \to \underline{Y}$ for each map~$f : X_1 \times
    \cdots \times X_n \to Y$,
    \\[-1.2em]
    \item one~$n$-ary relation symbol~$\underline{R} \hookrightarrow
    \underline{X_1} \times \cdots \times \underline{X_n}$ for each relation~$R
    \subseteq X_1 \times \cdots \times X_n$, and
    \\[-1.2em]
    \item an additional unary relation symbol~$G \hookrightarrow \underline{L}$
    (for the \emph{generic filter} of~$L$),
  \end{enumerate}
  we recursively define:}
  \scriptsize
  \begin{tabbing}
    \quad \= $\sigma \forces \forces \forall(x\?\underline{X})\_ \varphi$ \=
    \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\
    \forall(x_0 \in X)\_ \tau \forces \varphi[\underline{x_0}/x]$.\qquad\quad \=
    $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
    \= $\sigma \forces \underline{R}(\underline{s_1},\ldots,\underline{s_n})$ \= \defeqvi $s = t$. \= \kill

@@

    \> $\sigma \forces \forall(x\?\underline{X})\_ \varphi$
    \> \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\ \forall(x_0 \in X)\_ \tau \forces
    \varphi[\underline{x_0}/x]$.
    \> $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
    \> \defeqvi $\hil{$\nabla\sigma\_$}\ \exists(x_0 \in X)\_ \sigma \forces \varphi[\underline{x_0}/x]$.
  \end{tabbing}
  \small
  \only<1>{Finally, we say that~$\varphi$ ``holds in~$V^\nabla$'' iff for all~$\sigma
  \in L$, $\sigma \forces \varphi$.}

  \footnotesize
  \begin{tabular}{@{}lp{0.27\textwidth}p{0.48\textwidth}}
    \toprule
    forcing notion & statement about~$V^\nabla$ & external meaning \\
    \midrule
    surjection $\NN \twoheadrightarrow X$ &
    ``the gen.\@ surj.\@ is surjective'' &
    $\forall(a{\in}X)\_ \forall(\sigma{\in}X^*)\_ \nabla(\tau{\preccurlyeq}\sigma)\_ \exists(n{\in}\NN)\_ \tau[n] = a$. \\[0.5em]
    \pause

    map $\NN \to X$ &
    ``the gen.\@ sequence is good'' &
    $\mathsf{Good} \mid [\,]$. \\[0.5em]

    frame of opens &
    ``every complex number has a square root'' &
    For every open~$U \subseteq X$ and every cont.\@
    function $f : U \to \CC$, there is an open covering $U = \bigcup_i U_i$ such
    that for each index~$i$, there is a cont.\@ function $g : U_i \to \CC$
    such that~$g^2 = f$. \\[4.5em]

    Zariski &
    ``$x \neq 0 \Rightarrow \text{$x$ inv.}$'' &
    if the only f.p.\@ $k$-algebra in which~$x = 0$ is the zero algebra,
    then~$x$ is invertible in~$k$
  \end{tabular}
\end{frame}

@@

\newcommand{\Good}{\mathsf{Good}}
\newcommand{\vinfer}[2]{\left.#1\quad\middle|\quad\begin{array}{l}#2\end{array}\right.}
\begin{frame}[plain]
  \scriptsize
  \[
    \renewcommand{\arraystretch}{1.3}
    \vinfer{\Good \mid [\,]}{
      \vinfer{\Good \mid 0}{
        \vinfer{\Good \mid 00}{\Good\ 00} \\[1em]
        \vinfer{\Good \mid 01}{\Good\ 01} \\[1em]
        \vinfer{\Good \mid 02}{\Good\ 02} \\
        \vdots
      } \\[1em]
      \ \\[0.5em]
      \vinfer{\Good \mid 1}{
        \vinfer{\Good \mid 10}{
          \vinfer{\Good \mid 100}{\Good\ 100} \\[1em]
          \vinfer{\Good \mid 101}{\Good\ 101} \\[1em]
          \vinfer{\Good \mid 102}{\Good\ 102} \\
          \vdots
        } \\[1em]
        \vinfer{\Good \mid 11}{\Good\ 11} \\[1em]
        \vinfer{\Good \mid 12}{\Good\ 12} \\[1em]
        \vdots
      } \\[1em]
      \ \\[0.5em]
      \vinfer{\Good \mid 2}{} \\[1em]
      \vdots
    }
  \]
\end{frame}
